Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-23 paper/introduction.tex
meg

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g. words that contain "z" are rare) \cite{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc, e.g. \cite{huth2016natural}), and investigate the resulting parameters of the best model.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \cite{cichy2014resolving, king2016brain, kriegeskorte2008representational}.

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity as this factor covaries with numerous other factors such as word length (short words are more frequent than long words) word category (determinants are more frequent than adverbs), individual letters etc (e.g. words that contain "z" are rare) \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc, e.g. \citep{huth2016natural}), and investigate the resulting parameters of the best model.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational}.

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \citep{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.

### 2019-05-23 paper/meg_exps.tex
meg

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject. The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram. The diverging curves correspond to artifacts in data collection.}
  \label{fig:megavg}
\end{figure}

\subsubsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
\citep{spacy2} and WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency).  This
procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
$d=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We compare the encoding approach (columns 1 and 2, ridge regression $X \rightarrow Y$) and the decoding approach (column 3, ridge regression $Y \rightarrow X$) to our method (column 4, $X \rightarrow Y \rightarrow X$) on real MEG data from the MOUS dataset. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. Our method can disambiguate them consistently with
  previous neuroscientific work showing an early response ($\approx$ 150ms) to
  length and a later response ($\approx$ 400ms) to frequency.}
  \label{fig:meg_twocurves}
\end{figure}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject. The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram. The diverging curves correspond to artifacts in the MEG data.}
  \label{fig:megavg}
\end{figure}

\subsubsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using Spacy \citep{spacy2} and
WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency). This
procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
$d=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We applied the forward regression (a.k.a encoding coefficients), the backward regression (a.k.a decoding scores) and B2B on real MEG data. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. Our method can disambiguate them consistently with
  previous neuroscientific work showing an early response ($\approx$ 150ms) to
  length and a later response ($\approx$ 400ms) to frequency.}
  \label{fig:meg_twocurves}
\end{figure}
